Nock for Everyday Coders

### 2025-07-03 mss.tex
Switch to loobean

\begin{abstract}
  Nock is not super complex, and most programmers can learn the basics of it rapidly. The mental model gained by learning Nock turns out to be very useful in learning Hoon and understanding Urbit.
  While the Urbit docs generally suggest to not worry about Nock, Nock is very simple and small.  Most programmers will feel more comfortable in Hoon having learned Nock's basics.
  The goal of this tutorial is to explain Nock clearly in terms most programmers will relate to, to impart a feeling of confidence with very basic Nock, and to give a knowledge of Nock’s idioms and big ``wins" so that they carry over to learning Hoon.

@@

\begin{itemize}
  \item  \lstinline[style=inlinecode]{?}: ``check whether the value after this is a cell. Return 0 if yes, 1 if no".
  \item  \lstinline[style=inlinecode]{=}: ``first run the function in \lstinline[style=inlinecode]{b} with subject \lstinline[style=inlinecode]{a} as the argument, and same for the function in \lstinline[style=inlinecode]{c}. If the results are equal, return 0, if not, return 1.
\end{itemize}

@@

This is exactly what we want: the program crashes unless we are doing a boolean test that returns a \lstinline[style=inlinecode]{0} or \lstinline[style=inlinecode]{1} (converted to an indicial \lstinline[style=inlinecode]{2} or \lstinline[style=inlinecode]{3}) in step 4.

@@

Now I'm going to make you a little sad. Most Nock interpreters don't do this whole awesome code expansion. They just see \lstinline[style=inlinecode]{@6@}, and implement an if/else check with a crash if \lstinline[style=inlinecode]{*[a b]} isn't a boolean.

@@

To finish this off and take your new powers for a spin, let's look at some real Nock code from the wild. I got the below example from the Dojo. It's a mold: a function that takes a noun and returns it if it's the correct type, and crashes if not. The mold here checks whether the input noun is a boolean (\lstinline[style=inlinecode]{0}/true or \lstinline[style=inlinecode]{1}/false).

\subsubsection{Example: A Boolean Mold}

\begin{lstlisting}[style=listingcode]
> =boolean-mold ?

:: below output shortened for our purposes here
> boolean-mold
< 1.toz ... >

::  grab the code for the battery
> -.boolean-mold
[@6@ [@5@ [@1@ 0] @0@ 6] [@1@ 0] @6@ [@5@ [@1@ 1] @0@ 6] [@1@ 1] @0@ 0]
\end{lstlisting}

@@

\begin{enumerate}
  \item  Assign the mold gate denoted by \lstinline[style=inlinecode]{?} to the face \lstinline[style=inlinecode]{boolean-mold}.
  \item  Examine what's inside, seeing that the head is an arm.
  \item  Print out that source code by calling the head.
\end{enumerate}

\noindent{}
We know the below code is a gate that evaluates its sample, and returns it if it's a boolean, and crashes otherwise. Let's see how that works.

@@

We know the sample will be a noun that we're testing for booleanness, so this code starts by seeing whether the sample is the value \lstinline[style=inlinecode]{0}/true. If it is, the next element is \lstinline[style=inlinecode]{[@1@ 0]}, so we return \lstinline[style=inlinecode]{0} if the sample is \lstinline[style=inlinecode]{0}.

@@

If not, it means our input was neither \lstinline[style=inlinecode]{0} nor \lstinline[style=inlinecode]{1} and is not a boolean, so we run the formula \lstinline[style=inlinecode]{[@0@ 0]}, which always crashes (as there is no memory slot \lstinline[style=inlinecode]{0}). This is exactly what we want—crash if the sample is not a boolean!

@@

\begin{lstlisting}[style=listingcode]
++  my-function
  ~/  %my-function
  |=  var=@
  ...
\end{lstlisting}

@@

\subsection{Nock in Hoon}.

>>>>

\begin{abstract}
  \sloppy
  Nock is not super complex, and most programmers can learn the basics of it rapidly. The mental model gained by learning Nock turns out to be very useful in learning Hoon and understanding \mbox{Urbit}.  While the Urbit docs generally suggest to not worry about Nock, Nock is very simple and small.  Most programmers will feel more comfortable in Hoon having learned Nock's basics.  The goal of this tutorial is to explain Nock clearly in terms most programmers will relate to, to impart a feeling of confidence with very basic Nock, and to give a knowledge of Nock’s idioms and big wins so that they carry over to learning Hoon.

@@

\begin{itemize}
  \item  \lstinline[style=inlinecode]{?}: ``check whether the value after this is a cell. Return 0 if yes, 1 if no".\footnote{Nock conventionally explains that there is one ``true'' case and (potentially) many ``false'' cases, thus 0/\texttt{true} and 1/\texttt{false}.}
  \item  \lstinline[style=inlinecode]{=}: ``first run the function in \lstinline[style=inlinecode]{b} with subject \lstinline[style=inlinecode]{a} as the argument, and same for the function in \lstinline[style=inlinecode]{c}. If the results are equal, return 0, if not, return 1.
\end{itemize}

@@

This is exactly what we want: the program crashes unless we are doing a loobean\footnote{A boolean truth value but with the typical truth values of 0 and 1 reversed to 0/\texttt{true} and 1/\texttt{false}.} test that returns a \lstinline[style=inlinecode]{0} or \lstinline[style=inlinecode]{1} (converted to an indicial \lstinline[style=inlinecode]{2} or \lstinline[style=inlinecode]{3}) in step 4.

@@

Now I'm going to make you a little sad. Most Nock interpreters don't do this whole awesome code expansion. They just see \lstinline[style=inlinecode]{@6@}, and implement an if/else check with a crash if \lstinline[style=inlinecode]{*[a b]} isn't a loobean.

@@

To finish this off and take your new powers for a spin, let's look at some real Nock code from the wild. I got the below example from the Dojo. It's a mold: a function that takes a noun and returns it if it's the correct type, and crashes if not. The mold here checks whether the input noun is a loobean (\lstinline[style=inlinecode]{0}/true or \lstinline[style=inlinecode]{1}/false).

\subsubsection{Example: A Loobean Mold}

\begin{lstlisting}[style=listingcode]
:: loobean mold gate
> =loobean-mold ?

:: below output truncated for our purposes here
> loobean-mold
< 1.fxf ... >

::  grab the code for the battery
> -.loobean-mold
[@8@ [@6@ [@5@ [@1@ 0] @0@ 6] [@1@ 0] @6@ [@5@ [@1@ 1] @0@ 6] [@1@ 1] @0@ 0]
 @8@ [@5@ [@0@ 14] @0@ 2] @0@ 6]
\end{lstlisting}

@@

\begin{enumerate}
  \item  Assign the mold gate denoted by \lstinline[style=inlinecode]{?} to the face \lstinline[style=inlinecode]{loobean-mold}.
  \item  Examine what's inside, seeing that the head is an arm.
  \item  Print out that source code by calling the head.
\end{enumerate}

\noindent{}
We know the below code is a gate that evaluates its sample, and returns it if it's a loobean, and crashes otherwise. Let's see how that works.

@@

We know the sample will be a noun that we're testing for loobeanness, so this code starts by seeing whether the sample is the value \lstinline[style=inlinecode]{0}/true. If it is, the next element is \lstinline[style=inlinecode]{[@1@ 0]}, so we return \lstinline[style=inlinecode]{0} if the sample is \lstinline[style=inlinecode]{0}.

@@

If not, it means our input was neither \lstinline[style=inlinecode]{0} nor \lstinline[style=inlinecode]{1} and is not a loobean, so we run the formula \lstinline[style=inlinecode]{[@0@ 0]}, which always crashes (as there is no memory slot \lstinline[style=inlinecode]{0}).\footnote{In the binary tree addressing scheme of Nock, 1 refers to the entire noun, 2 to its head if it exists, and 3 to its tail if it exists.} This is exactly what we want—crash if the sample is not a loobean!

@@

\begin{lstlisting}[style=listingcode]
++  my-function
  ~/  %my-function
  |=  var=*
  ...
\end{lstlisting}

@@

\subsection{Nock in Hoon}
